---
title: "COVID19"
author: "J. Achalapong"
date: "2025-07-19"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Step1: Import Data

This section initiates the project by importing the necessary packages and reading in four datasets from the Johns Hopkins University CSSE COVID-19 GitHub repository. The datasets include global and US COVID-19 confirmed cases and deaths. The data is loaded into R using read_csv() for further processing.  


```{r get_jhu_data}
library(tidyverse)
library(lubridate)

url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/refs/heads/master/csse_covid_19_data/csse_covid_19_time_series/"

file_names <- c("time_series_covid19_confirmed_global.csv",
                "time_series_covid19_deaths_global.csv",
                "time_series_covid19_confirmed_US.csv",
                "time_series_covid19_deaths_US.csv")

urls <- str_c(url_in,file_names)
urls
```

Let's read in the data and see what we have.
```{r import_data, message=FALSE}
global_cases <- read_csv(urls[1])
global_deaths <- read_csv(urls[2])
us_cases <- read_csv(urls[3])
us_deaths <- read_csv(urls[4])
```

## Step2: Tidy and Transform Data

The raw datasets are in a wide format with date columns spread across. To prepare for analysis, the data is reshaped into a long format using `pivot_longer()`. Columns such as latitude and longitude are removed, and death and case data are joined together. `Dates` are parsed into Date objects. For US data, the same transformations are applied, and a `Combined_Key` is used to merge data with a population lookup table to facilitate normalization of case and death counts.  


```{r tidy_global_data}
global_cases <- global_cases %>%
  pivot_longer(cols = -c(`Province/State`,
                         `Country/Region`, Lat, Long),
               names_to = "date",
               values_to = "cases") %>%
  select(-c(Lat,Long))

global_deaths <- global_deaths %>%
  pivot_longer(cols = -c(`Province/State`,
                         `Country/Region`, Lat, Long),
               names_to = "date",
               values_to = "deaths") %>%
  select(-c(Lat,Long))

global <- global_cases %>%
  full_join(global_deaths) %>%
  rename(Country_Region = `Country/Region`,
         Province_State = `Province/State`) %>%
  mutate(date = mdy(date))

summary(global)

global <- global %>% filter(cases > 0)
```

```{r tidy_us_data}
us_cases <- us_cases %>%
  pivot_longer(cols = -c(UID:Combined_Key),
               names_to = "date",
               values_to = "cases") %>%
  select(Admin2:cases) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

us_deaths <- us_deaths %>%
  pivot_longer(cols = -c(UID:Population),
               names_to = "date",
               values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

us <- us_cases %>%
  full_join(us_deaths)

global <- global %>%
  unite("Combined_Key",
        c(Province_State, Country_Region),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE)

uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/refs/heads/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"

uid <- read_csv(uid_lookup_url) %>% select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))

global <- global %>%
  left_join(uid, by = c("Province_State", "Country_Region")) %>%
  select(-c(UID, FIPS)) %>%
  select(Province_State, Country_Region, date,
         cases, deaths, Population,
         Combined_Key)

global
```
## Step3: Visualizing Data

This section uses `ggplot2` to plot the number of cases and deaths over time in the United States, both at the national and state levels (e.g., New York). Logarithmic scaling is applied to the y-axis to handle the wide range of case and death counts. The graphs help reveal trends and highlight spikes in the pandemic timeline.  


``` {r visualizing_data}
us_by_state <- us %>%
  group_by(Province_State, Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(death_per_mill = deaths *1000000 / Population) %>%
  select(Province_State, Country_Region, date,
         cases, deaths, death_per_mill, Population) %>%
  ungroup()

us_total <- us_by_state %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(death_per_mill = deaths *1000000 / Population) %>%
  select(Country_Region, date,
         cases, deaths, death_per_mill, Population) %>%
  ungroup()

us_total %>%
  filter(cases > 0 ) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(colour = "cases")) +
  geom_point(aes(colour = "cases")) +
  geom_line(aes(y = deaths, colour = "deaths")) +
  geom_point(aes(y = deaths, colour = "deaths")) +
  scale_y_log10()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID19 in US", y = NULL)

state <- "New York"
us_by_state %>%
  filter(Province_State == state) %>%
  filter(cases > 0 ) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(colour = "cases")) +
  geom_point(aes(colour = "cases")) +
  geom_line(aes(y = deaths, colour = "deaths")) +
  geom_point(aes(y = deaths, colour = "deaths")) +
  scale_y_log10()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("COVID19 in ", state), y = NULL)

state <- "Colorado"
us_by_state %>%
  filter(Province_State == state) %>%
  filter(cases > 0 ) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(colour = "cases")) +
  geom_point(aes(colour = "cases")) +
  geom_line(aes(y = deaths, colour = "deaths")) +
  geom_point(aes(y = deaths, colour = "deaths")) +
  scale_y_log10()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("COVID19 in ", state), y = NULL)
```

## Step4: Analyzing Data
New daily case and death counts are computed using the `lag()` function. These derived variables (`new_cases`, `new_deaths`) are plotted to visualize the progression and waves of the pandemic. Warnings indicate missing or infinite values due to log-scaling and differences involving zeros.  


``` {r analyzing_data}
us_by_state <- us_by_state %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))

us_total <- us_total %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))

us_total %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(colour = "new_cases")) +
  geom_point(aes(colour = "new_cases")) +
  geom_line(aes(y = new_deaths, colour = "new_deaths")) +
  geom_point(aes(y = new_deaths, colour = "new_deaths")) +
  scale_y_log10()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID19 in US", y = NULL)

state <- "New York"
us_by_state %>%
  filter(Province_State == state) %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(colour = "new_cases")) +
  geom_point(aes(colour = "new_cases")) +
  geom_line(aes(y = new_deaths, colour = "new_deaths")) +
  geom_point(aes(y = new_deaths, colour = "new_deaths")) +
  scale_y_log10()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("COVID19 in ", state), y = NULL)

state <- "Colorado"
us_by_state %>%
  filter(Province_State == state) %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(colour = "new_cases")) +
  geom_point(aes(colour = "new_cases")) +
  geom_line(aes(y = new_deaths, colour = "new_deaths")) +
  geom_point(aes(y = new_deaths, colour = "new_deaths")) +
  scale_y_log10()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("COVID19 in ", state), y = NULL)

us_state_totals <- us_by_state %>%
  group_by(Province_State) %>%
  summarize(deaths = max(deaths), cases = max(cases),
            Population = max(Population),
            cases_per_thou = 1000* cases / Population,
            deaths_per_thou = 1000* deaths / Population) %>%
  filter(cases > 0, Population > 0)

us_state_totals %>%
  slice_min(deaths_per_thou, n=10) %>%
  select(deaths_per_thou, cases_per_thou, everything())

us_state_totals %>%
  slice_max(deaths_per_thou, n=10) %>%
  select(deaths_per_thou, cases_per_thou, everything())
```

## Step5: Modeling Data

A simple linear regression is conducted to model the relationship between cases per thousand (`cases_per_thou`) and deaths per thousand (`deaths_per_thou`) using `lm()`. The model shows a statistically significant positive relationship, suggesting that states with more cases per capita tend to have more deaths per capita. The model’s predictions are visualized alongside the actual data points.  


```{r modeling_data}
mod <- lm(deaths_per_thou ~ cases_per_thou, data = us_state_totals)
summary(mod)

us_state_totals %>% slice_min(cases_per_thou)
us_state_totals %>% slice_max(cases_per_thou)

x_grid <- seq(1,451)
new_df <- tibble(cases_per_thou = x_grid)
us_state_totals %>% mutate(pred = predict(mod))

us_tot_w_pred <- us_state_totals %>% mutate(pred = predict(mod))
us_tot_w_pred %>% ggplot() +
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") +
  geom_point(aes(x = cases_per_thou, y = pred), color = "red")
```



## Step 6: Report Conclusion and Sources of Bias

This analysis provides a high-level view of the COVID-19 pandemic in the United States and globally, offering insights into trends, state-wise impacts, and the relationship between infection and mortality rates.  

### Key conclusions  

 - The progression of cases and deaths followed a pattern of waves.  
 - Some states experienced much higher case and death rates per capita.  
 - There is a significant linear relationship between case incidence and death incidence.  

### Potential sources of bias and limitations  

 - Underreporting or inconsistent testing: COVID-19 case counts depend heavily on testing availability and public health reporting systems.  

 - Population data limitations: Mismatches between case/death data and population data could introduce bias in per-capita calculations.  

 - Data lag and reporting delays: These affect the accuracy of daily case and death counts, especially visible in the spikes and drops.  

 - Policy differences: Differences in public health measures, reporting standards, and healthcare infrastructure across states may confound the observed relationships.  

 - Simplistic modeling: The linear regression used does not account for other influential factors such as age demographics, vaccination rates, or comorbidities.  
 